Supplement to Automatic Matching of Bullet Lands

### 2016-01-20 imaging-supplement.Rnw
proof-read done for supplement

\section{Cylindrical Fit}\label{supp:cylindrical}
Figure~\ref{fig:fixedX} shows the side profile of a crosscut through the surface measurements of bullet 1-5 at a fixed level. The smooth line on top is a circle, with estimated radius and center. The details of this fit are given below:
%
\begin{figure}[hbtp]
  \centering
<<fixedX, dependson='data', echo=FALSE, warning=FALSE, message=FALSE, fig.height=2, fig.width=6, out.width='0.5\\textwidth'>>=

@@

qplot(data=subset(dbr111, x <= 100*1.5625^2 & x >= 99*1.5625^2), y, value, geom="line", size=I(1)) +
  geom_line(aes(x=xpred, y=ypred, group=x), 
            colour="grey30", size=0.25) +
#  ylab(expression(paste("Surface Measurements (in ",mu,m,")", sep=""))) + 
  ylab("") +
  theme_bw() + 
  theme(legend.position="bottom") #+ coord_equal()
@
\caption{\label{fig:fixedX}Side profile of the surface measurements (in $\mu m$) of a bullet land across a fixed $x$ value. Note that the global features dominate any deviations, corresponding to the individual characteristics of striation marks.}
\end{figure}

@@

The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit.
In this instance, the radius is estimated as $\hat{r} = \Sexpr{pars$radius}\mu m = \Sexpr{pars$radius/1000}mm$ and the land covers about \Sexpr{diff(range(dbr111$theta))} degrees.  Both of these estimates are consistent with a 9 mm bullet fired by a Ruger P-85.
The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a surface cross cut at $x = 100$ there is a residual circular structure that does not show up for all signatures.

\begin{figure}[hbtp]
  \centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{\label{fig:residuala}Residual structure at $x = 1.5625$ (bottom of the bullet).}
<<residual2, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\textwidth'>>=
qplot(data=subset(dbr111, x <= 1.5625), y, value-ypred, #colour=factor(x),
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  theme_bw() + 
  geom_hline(yintercept = 0, colour="grey50") +
  ylab(expression(paste("Residuals (in ",mu,"m)", sep=""))) + 
  theme(legend.position="bottom")
@
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{\label{fig:residualb} Residual structure at $x = 100.00$}
<<residual, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\textwidth'>>=
#qplot(data=subset(dbr111, x <= 80*1.5625^2 & x >=75*1.5625^2), y, value-ypred,
qplot(data=subset(dbr111, x == 100), y, value-ypred,  #colour=factor(x), 
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  geom_hline(yintercept = 0, colour="grey50") +
  theme_bw() + 
  ylab(expression(paste("Residuals (in ",mu,"m)", sep=""))) + 
  theme(legend.position="bottom")
@
\end{subfigure}
\caption{\label{fig:residual} Residual structure of circular fits at two different cross sections. Both residual plots show systematic structures, indicating that a circular fit is not entirely appropriate.}
\end{figure}

A single cylinder as a fit is unlikely to be a particularly good fit, because there seem to be quite massive deformations in the vertical direction. Even when we fit a circle at each cross section of the bullet, as in Figure~\ref{fig:circlefits}, this does not address all of these issues. While the wider circumference at the base of the bullet can be resolved by individual circular fits, the systematic residual structure in Figure~\ref{fig:residualb} stays the same.

@@

\newpage
\section{Assessing cross-correlation between signatures at multiple cross sections}\label{supp:ccf}

Figure~\ref{fig:crosscuts} shows a sequence of signatures for bullet 1-5 (barrel 1) that are taken at cross sections 50$\mu m$ apart, between 150$\mu m$  and 400$\mu m$. These are compared to the signature at cross section 100$\mu m$. Initially, this comparison constitutes an almost perfect match between the two signatures. However, the match quickly deteriorates with increasing distance between the cross sections.  Only if signatures are from cross sections  within 150$\mu m$ do we get a good visual match even when we know that the same bullet surface is being used. 
Given that we have to expect some variation in nominally the same cross section levels due to manual alignments in microscopes, we should take cross section levels into account in the automatic matching routine by evaluating matches at several cross sections. 
\begin{figure}[hbtp]
  \centering
<<crosscuts-vary, echo=FALSE, fig.width = 12, fig.height = 7, out.width = '\\textwidth', cache=TRUE, warning = FALSE>>=

@@

grid.arrange(lop[[1]], lop[[2]], lop[[3]], lop[[4]], lop[[5]], lop[[6]],
             ncol = 2)
@
\caption{\label{fig:crosscuts}Overview of the variations in the surface along different cross sections. The signature extracted at $x = 100$ is compared to signatures at every 50$\mu m$. With every step away from the original cross section site, the number of differences between the signatures increases; the number of maximum CMS decreases from initially 22 to  four or fewer at $x = 300\mu m$ and higher. }
\end{figure}

@@

  write.csv(stats, "csvs/crosscut-sd.csv", row.names=FALSE)
} else {
  stats <- read.csv("csvs/crosscut-sd.csv")
}  
@
%
Figure~\ref{fig:overview} shows an overview of the signatures across different crossucts on a single bullet. 
\begin{figure}[hbtp]
<<one-bulletland-sd, echo=FALSE, dependson='setup-signatures', fig.height=5, fig.width=8, out.width='0.7\\textwidth', warning=FALSE>>=
k <- 1
x <- as.character(ccs$path)[k]
land <- read.x3p(x)
land[[3]] <- x
names(land)[3] <- "path"

@@

bullet <- dframe %>% bind_rows %>% bulletSmooth 
subbullet <- subset(bullet, x >= ccs$cc[k])
qplot(x=y, geom="line", y=l30, colour=x, group=x, data=subbullet) + theme_bw() +
  scale_colour_gradient("Crosscuts") +
  theme(legend.position="bottom") +
  ylab("Residuals from loess fit")
@
\caption{\label{fig:overview}Signatures of the same bullet at different crosscuts. With increasing crosscut level, peaks and valleys are less pronounced, resulting in a smaller standard deviation.}
\end{figure}
As crosscut levels increase individual characteristics become less distinctive, making true matches to other bullets harder. The pattern of decreasing peaks and valleys is generally true for  bullet lands, as can be seen in Figure~\ref{fig:sds}. 
%
\begin{figure}[hbtp]
\centering
<<crosscuts-sd, echo=FALSE, fig.height=5, fig.width=8, out.width='.7\\textwidth', dependson='setup-signatures'>>=
qplot(x, sd, data=stats, geom="line", group=bullet, alpha=I(0.5)) + theme_bw() +
  geom_smooth(group=1) + 
  xlab("Crosscut location") +
  ylab("Standard deviation")
@
\caption{\label{fig:sds}Standard deviation reduces as crosscut levels increase.}
\end{figure}
Figure~\ref{fig:sds} shows that the amount of standard deviation of a signature decreases on average for all bullet lands as crosscut levels increase.
This makes standard deviation of a signature one measure to quantify the extent to which a signature is expressed. This implies that for identifying matches we should use the lowest crosscut level to extract a bullet's signature once a stable surface region is detected. This is in accordance with current standard practice \citep{afte:1992}.

%\newpage
\section{Complete evaluation of the Hamby study}\label{supp:extended}
\hh{One way to expand the use of the James Hamby study is to not only match all of the unknown bullet lands against the known bullet lands, but to compare every land against every other land. This effectively doubles the number of comparisons from 10,384 pairwise comparisons of usable bullet lands to 21,115 $\left[= (118+88)\cdot 205/2\right]$ comparisons by adding another 10,731 bullet land comparisons made up of known-to-known and unknown-to-unkown comparisons. }

\hh{When we predict the new 10,731 comparisons using the random forest based on the previously considered 10,384 known-unknown comparisons, we encounter 18 false negatives and 9 false positives, corresponding to an actual false error rate of 0.19 and a false positive rate of 0.00085, which is close to the random forest's estimated OOB error rates of 0.226744 and 0.000098. }

\hh{However, if we use all of the available comparisons to fit another random forest of 300 trees, the de-facto error rates for false positives and false negatives are again at 0. The estimated OOB error rates are 0.00024 for the false positive rate and 0.22180 for the false negative rate. The false positive rate is therefore virtually unchanged, while we see a slight improvement in the false negative rate for an overall OOB error rate of 0.3\%, i.e.\ an increase to twice the number of comparisons leads to a decrease of 25\% of the estimated error rate. This is yet another argument in favor of a larger database for training algorithms.}

\hh{XXX maybe use log of D to remove the left boundary XXX}
<<allhamby, echo=FALSE, message=FALSE, warning=FALSE>>=
knowndatadir <- "app/images/Hamby252_3DX3P1of2/"
knowns <- dir(path=knowndatadir)
knowns <- gsub(".x3p", "", knowns)

>>>>

\section{Cylindrical Fit}\label{supp:cylindrical}
Figure~\ref{fig:fixedX} shows the  profile of surface measurements of bullet 1-5 at a fixed height. The smooth line on top is a circle, with estimated radius and center. The details of this fit are given below:
%
\begin{figure}[hbtp]
  \centering
<<fixedX, dependson='data', echo=FALSE, warning=FALSE, message=FALSE, fig.height=2, fig.width=6, out.width='0.5\\textwidth'>>=

@@

qplot(data=subset(dbr111, x <= 100*1.5625^2 & x >= 99*1.5625^2), y, value, geom="line", size=I(1)) +
  geom_line(aes(x=xpred, y=ypred, group=x), 
            colour="grey30", size=0.25) +
#  ylab(expression(paste("Surface Measurements (in ",mu,m,")", sep=""))) + 
  ylab("") +
  theme_bw() + 
  theme(legend.position="bottom") #+ coord_equal()
@
\caption{\label{fig:fixedX}Side profile of the surface measurements (in $\mu m$) of a bullet land at a fixed height of $x$. Note that the global features dominate any deviations, corresponding to the individual characteristics of striation marks.}
\end{figure}

@@

The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit.
In this instance, the radius is estimated as $\hat{r} = \Sexpr{pars$radius}\mu m = \Sexpr{pars$radius/1000}mm$ and the land covers about \Sexpr{diff(range(dbr111$theta))} degrees.  Both of these estimates are consistent with a 9 mm bullet fired by a Ruger P-85.
The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a profile at height $x = 100\mu m$ there is a residual circular structure that does not show up for all signatures.

\begin{figure}[hbtp]
  \centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{\label{fig:residuala}Residual structure at height $x = 1.5625\mu m$ (bottom of the bullet).}
<<residual2, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\textwidth'>>=
qplot(data=subset(dbr111, x <= 1.5625), y, value-ypred, #colour=factor(x),
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  theme_bw() + 
  geom_hline(yintercept = 0, colour="grey50") +
  ylab(expression(paste("Residuals (in ",mu,"m)", sep=""))) + 
  theme(legend.position="bottom")
@
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{\label{fig:residualb} Residual structure at height $x = 100.00\mu m$}
<<residual, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\textwidth'>>=
#qplot(data=subset(dbr111, x <= 80*1.5625^2 & x >=75*1.5625^2), y, value-ypred,
qplot(data=subset(dbr111, x == 100), y, value-ypred,  #colour=factor(x), 
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  geom_hline(yintercept = 0, colour="grey50") +
  theme_bw() + 
  ylab(expression(paste("Residuals (in ",mu,"m)", sep=""))) + 
  theme(legend.position="bottom")
@
\end{subfigure}
\caption{\label{fig:residual} Residual structure of circular fits at two different cross sections. Both residual plots show systematic structures, indicating that a circular fit is not entirely appropriate.}
\end{figure}

A single cylinder as a fit is unlikely to be a particularly good fit, because there seem to be quite massive deformations in the vertical direction. Even when we fit a circle at each distinct height of the bullet, as in Figure~\ref{fig:circlefits}, this does not address all of these issues. While the wider circumference at the base of the bullet can be resolved by individual circular fits, the systematic residual structure in Figure~\ref{fig:residualb} stays the same.

@@

\newpage
\section{Assessing cross-correlation between signatures at multiple levels of height}\label{supp:ccf}

Figure~\ref{fig:crosscuts} shows a sequence of signatures for bullet 1-5 (barrel 1) that are taken at heights 50$\mu m$ apart, between 150$\mu m$  and 400$\mu m$. These are compared to the signature at a height of 100$\mu m$. Initially, this comparison constitutes an almost perfect match between the two signatures. However, the match quickly deteriorates with increasing distance between the heights at which signatures are extracted.  Only if signatures are from heights  within 150$\mu m$ do we get a good visual match even when we know that the same bullet surface is being used. 
Given that we have to expect some variation in nominally the same height values due to (manual) alignments in microscopes, we should take height values into account in the automatic matching routine by evaluating matches at several heights. 
\begin{figure}[hbtp]
  \centering
<<crosscuts-vary, echo=FALSE, fig.width = 12, fig.height = 7, out.width = '\\textwidth', cache=TRUE, warning = FALSE>>=

@@

grid.arrange(lop[[1]], lop[[2]], lop[[3]], lop[[4]], lop[[5]], lop[[6]],
             ncol = 2)
@
\caption{\label{fig:crosscuts}Overview of the variations in the signatures at different heights. The signature extracted at $x = 100\mu m$ is compared to signatures at every 50$\mu m$. With every step away from the original height, the number of differences between the signatures increases; the number of maximum CMS decreases from initially 22 to  four or fewer at a height of $x = 300\mu m$ and above. }
\end{figure}

@@

  write.csv(stats, "csvs/crosscut-sd.csv", row.names=FALSE)
} else {
  stats <- read.csv("csvs/crosscut-sd.csv")
}  
@
%
Figure~\ref{fig:overview} shows an overview of the signatures at different heights on a single bullet. 
\begin{figure}[hbtp]
<<one-bulletland-sd, echo=FALSE, dependson='setup-signatures', fig.height=5, fig.width=8, out.width='0.7\\textwidth', warning=FALSE>>=
k <- 1
x <- as.character(ccs$path)[k]
land <- read.x3p(x)
land[[3]] <- x
names(land)[3] <- "path"

@@

bullet <- dframe %>% bind_rows %>% bulletSmooth 
subbullet <- subset(bullet, x >= ccs$cc[k])
qplot(x=y, geom="line", y=l30, colour=x, group=x, data=subbullet) + theme_bw() +
  scale_colour_gradient("Crosscuts") +
  theme(legend.position="bottom") +
  ylab("Residuals from loess fit")
@
\caption{\label{fig:overview}Signatures of the same bullet at different heights.  With increasing height, peaks and valleys are less pronounced, resulting in a smaller standard deviation.}
\end{figure}
At larger heights  individual characteristics become less distinctive, making true matches to other bullets harder. The pattern of decreasing peaks and valleys is generally true for  bullet lands, as can be seen in Figure~\ref{fig:sds}. 
%
\begin{figure}[hbtp]
\centering
<<crosscuts-sd, echo=FALSE, fig.height=5, fig.width=8, out.width='.7\\textwidth', dependson='setup-signatures'>>=
qplot(x, sd, data=stats, geom="line", group=bullet, alpha=I(0.5)) + theme_bw() +
  geom_smooth(group=1) + 
  xlab("Height (from the bottom of the bullet)") +
  ylab("Standard deviation")
@
\caption{\label{fig:sds}Standard deviation reduces as height increase.}
\end{figure}
Figure~\ref{fig:sds} shows that the amount of standard deviation of a signature decreases on average for all bullet lands at larger heights.
This makes standard deviation of a signature one measure to quantify the extent to which a signature is expressed. For identifying matches we should therefore use the lowest height to extract a bullet's signature once a stable surface region is detected. This is in accordance with current standard practice \citep{afte:1992}.

%\newpage
\section{Complete evaluation of the Hamby study}\label{supp:extended}
One way to expand the use of the James Hamby study is to not only match all of the unknown bullet lands against the known bullet lands, but to compare every land against every other land. This effectively doubles the number of comparisons from 10,384 pairwise comparisons of usable bullet lands to 21,115 $\left[= (118+88)\cdot 205/2\right]$ comparisons by adding another 10,731 bullet land comparisons made up of known-to-known and unknown-to-unkown comparisons.

When we predict the new 10,731 comparisons using the random forest based on the previously considered 10,384 known-unknown comparisons, we encounter 18 false negatives and 9 false positives, corresponding to an actual false error rate of 0.19 and a false positive rate of 0.00085, which is close to the random forest's estimated OOB error rates of 0.226744 and 0.000098.

However, if we use all of the available comparisons to fit another random forest of 300 trees, the de-facto error rates for false positives and false negatives are again at 0. The estimated OOB error rates are 0.00024 for the false positive rate and 0.22180 for the false negative rate. The false positive rate is therefore virtually unchanged, while we see a slight improvement in the false negative rate for an overall OOB error rate of 0.3\%, i.e.\ an increase to twice the number of comparisons leads to a decrease of 25\% of the estimated error rate. This is yet another argument in favor of a larger database for training algorithms.

%\hh{XXX maybe use log of D to remove the left boundary XXX}
<<allhamby, echo=FALSE, message=FALSE, warning=FALSE>>=
knowndatadir <- "app/images/Hamby252_3DX3P1of2/"
knowns <- dir(path=knowndatadir)
knowns <- gsub(".x3p", "", knowns)
